\headingfont UNIVERSIDAD DE EXTREMADURA\\ FACULTAD DE CIENCIAS\\[1.5cm] \textbf{ Grado en Física\\ TRABAJO DE FIN DE GRADO\\[3.5cm]} {\setstretch{1.25}\flushleft{\large{\textbf{Developement of a FIWARE-based application for tree species monitoring (dendrometry)}}\\[3.5cm]}}

### 2020-07-08 docs/LaTex/Paper.tex
Final redaction fixes

\begin{itemize}
  \item \textbf{PoE} (Power-over-Ethernet); this is not actually a suitable option because Raspberry Pi does not support PoE by itself, it needs a \href{https://www.raspberrypi.org/products/poe-hat/}{separate HAT} using the GPIO connector, so this makes it impossible to use it along the LoRa HAT. Apparently there are hacks to reduce this HAT size, but they also use pins in the GPIO connector \cite{RaspiPoEHack}. Besides, this method only works with up to 100m cables, so it would be necessary to have a power source and internet access point within a 100m radius.\footnote{The \href{https://github.com/hallard/LoRasPI}{LoRasPI HAT} is more suitable for this option because does not cover whole GPIO connector, leaving free the GPIO pins used in \cite{RaspiPoEHack}.}
  \item \textbf{WiFi} and battery powered; this setup implies also the existence of a relatively close power source, because the access point used to provide that wireless network connection should work also with mains power. Probably the WiFi radio is not enough to make it the difference between this setup and the next one suggested.
  \item Simple \textbf{ethernet} and mains power, this setup is similar to the previous one, but omitting the WiFi limitations/configurations, the access point and the Raspberry Pi should be practically both at the same location.
  \item \textbf{GSM module} and battery or mains powered; this document considers this as the ideal setup. This setup does not require a conventional access point because the internet access is granted through a GSM module, in a similar way than a mobile phone. This would be the ideal way because it minimizes the resources needed, which is crucial in an environment where those probably will not be available ---in the middle of the forest.

@@

\begin{itemize}
  \item \textbf{IoTagent-LoRaWAN}, which is in charge of forwarding the traffic from TTN stack to Orion Context Broker.
  \item \textbf{Orion Context Broker}, this is the \myuline{core generic enabler}, in fact as indicated, this is the only one required component so the solution can be considered as \enquote{powered by FIWARE}. Besides, this generic enabler requires a \textbf{Mongo database} to store entities and subscriptions among other things. 
  \item \textbf{QuantumLeap}, in order to make the context data persistent, this generic enabler is needed. Orion Context Broker works mainly with two elements; entities and subscriptions, however, for any created entity the context data does not persist in Mongo database, instead the context data will be replaced/updated; so creating a subscription which notifies to QuantumLeap, this generic enabler will store the context data in a \textbf{Crate database} to make it persistent. Also a \textbf{Redis database} is used for geocoding purposes, however this project will not deal with it ---this is why of this database is not even being included in the figure \ref{fig:GenView}.
  \item \textbf{Grafana}, this container is intended to provide a web interface to visualize and manage the retrieved context data. It stores its configuration and customization values in an embebed sqlite3 database; so there is not need for a separate container. 
\end{itemize}

The terms above highlighted in bold are the deployed containers that will be detailed later, many of them are not even part of FIWARE platform, though. First this document will expose how to set up the gateway, create an application and add devices inside The Things Network stack.

@@

\subsubsection{Grafana}\label{sssec:Grafana}
Grafana is a powerful dashboard used at many scenarios in the software industry, it allows monitoring almost everything the reader can imagine; because of this, it is capable of acquiring the data from multiple sources. In this project it will be used in order to monitor the context data along the time.

@@

To get the panel plotting the context data stored in CrateDB, the reader must specify a query to retrieve the data from the database. This can be done in the \enquote{Query} tab below the graph shown in picture \ref{fig:panel_setup}: here the user could use the graphical interface or write the query manually by clicking on the pencil button at the right, so essentially the query should look as follows in example \ref{ej:CrateDB_query}

@@

Grafana has a wide options set intended to monitor different kind of systems, it can plot data in a variety of graphics and it even allows exporting to csv format, this can be useful for a more deep analysis using other technologies like, for instance, R, a well-known language widely used for statistical analysis.

Of course, this context data could have been acquired directly from CrateDB through some custom application developed specifically for this task, however this is beyond the purpose of this project. Perhaps in a future, a feature like this could be implemented.

@@

When the requirements are satisfied and all docker containers are properly running\footnote{In a first place the GUI will retrieve the versions for Orion Context Broker, IoTagent-LoRaWAN and QuantumLeap, so if they are not running the GUI will not open.} the GUI can be run performing \cmd{./api-gui.py}; this should open the window shown by figure \ref{fig:GUI_app_keys}

@@

To check Orion Context Broker, IoTagent-LoRaWAN and QuantumLeap versions running in the docker containers the reader has to click on \enquote{About} tab, however, this first shown tab ---\enquote{Application Keys}--- is the most important one because without completing all the required info the \enquote{Devices Management} tab will not be enabled. Actually this tab performs a series of checks regarding TTN keys. Following the same TTN requirements,

@@

When the keys are updated properly the labels over the text fields are also updated with the chosen values and the \enquote{Devices Management} tab is enabled. This tab provides a user-friendly interface to manage devices. In a similar way, this info will must be the same than the info provided by TTN interface, so as in figure \ref{fig:device_management_tab}

@@

This view allows the user to create and delete devices ---along with its entities and subscriptions--- nevertheless, the \enquote{Delete device and entities} button remains disabled until a device is chosen in the right list. In a similar way than \enquote{Applications Keys}, to add a new device to the FIWARE instance deployed by docker, all keys must fullfil a set of minimum requirements; as it can be seen in picture \ref{fig:adding_dev}

\begin{itemize}
  \item Device entity and ID must be also unique and it has to be at least 3 characters long. In fact, this ID must match the chosen ID in TTN interface when the device is created,
  \item Application Session Key must be 32 characters long and also an hexadecimal string.
  \item Device EUI has to be 16 characters long and again an hexadecimal string.
\end{itemize}

These keys can be obtained from TTN as it can be seen in screenshot \ref{fig:dev_overview} ---this screenshot shows info about \cmd{node\_0}, though. We can see in picture \ref{fig:adding_dev}, that there was a problem with \enquote{Device EUI}, this was caused because the final \cmd{Z} letter in the string does not belong to hexadecimal characters set.

@@

Finally, as indicated, the \enquote{About} tab shows the versions info as it can be seen in \ref{fig:about_tab}. The QuantumLeap version is specified in \cmd{docker-compose.yml} file as well as the CrateDB version, this is to ensure compatibility between QuantumLeap and CrateDB; however, Orion Context Broker and IoTagent-LoRaWAN versions are not specified so they should be always the latests, this is why retrieving them could be helpful.

@@

\clearpage{\pagestyle{empty}\cleardoublepage}\thispagestyle{plain}  % Forces the next section to start in an odd page.
\section{Proof of concept, deployment}
Due to inconvenient circumstances at early 2020, this section is not as long as initially planned, instead of testing the system in a real production environment, this section will focus on summarizing needed steps/stages to properly deploy the system. Although these steps are already separately detailed, this section tries to put them together to give a consistent and practical point of view.

@@

\subsection{Set up the gateway}
This process has been widely detailed in \ref{sssec:Gateway} subsection. The most important point on this stage is to ensure the systemd service is working properly and to retrieve the Gateway ID, which will be needed in the next stage.

\subsection{TTN Stack}
Then, the reader should follow the three steps described in subsection \ref{sssec:TTN_Stack}: register a Gateway using its \enquote{Gateway ID}, add an application and register a device for that application. This is the first step because the reader will need the security keys for the two following stages:

\begin{itemize}
  \item Embed these keys in Arduino sketch. This sketch will require the \enquote{Application Session Key}, \enquote{Network Session Key} and \enquote{Device Address}.
  \item Use them to create devices, and subscriptions in FIWARE. To create the device ---and also its corresponding entity and subscription when using the designed graphical interface shown in \ref{sssec:API_GUI} subsection--- the IoTagent will require the \enquote{Application ID}, \enquote{Application Access Key}, \enquote{Device ID}, \enquote{Application EUI} and again \enquote{Application Session Key}.
\end{itemize}

We must remember also that it is very important to choose the CayenneLPP as the payload decoder. So if these keys are available in TTN platform, the process can continue.

\subsection{Customize the Arduino sketch for a specific device}
As indicated, this sketch will need the \enquote{Application Session Key}, the \enquote{Network Session Key} and \textbf{\enquote{Device Address}}, this latter is the most important one, because several nodes can use the same \enquote{Application Session Key} and \enquote{Network Session Key}, but \enquote{Device Address} must be unique ---it is in fact also provided by TTN and the reader can check that every newly created node has a different one in TTN platform.

Of course, to be able to compile the sketch some dependencies must be satisfied, such as CayenneLPP, ArduinoJson and lmic-arduino libraries. Moreover, the \textbf{linear potentiometer has been replaced for a rotary potentiometer}, again due to circumstances in the moment when the project is developed.

@@

It is important to note that analog pin 2 has been chosen the for analog signal acquisition. Besides, the reader can check the status of the device in the serial monitor, however, the reader may also note that serial monitor is setup at 115200 bits per second.

\subsection{Deploy Docker containers}
Once the hardware has been properly setup, the next step is to deploy locally all Docker containers, as indicated in \ref{sssec:FIWARE} subsection. Using \cmd{docker-compose} this becomes an easy task.

This will deploy essentially the IoTagent-LoRaWAN, Orion CB, QuantumLeap and Grafana. The status of these containers can be checked performing \cmd{docker ps -a}.

\subsection{Devices, entities and subscriptions}
When Docker containers are deployed, it should be possible to send HTTP requests to create a device (along its entity) and a subscription to notify QuantumLeap any change in any entity. This process can be done through two different ways as indicated previously,

\begin{itemize}
  \item Using developed graphical user interface (recommended, see \ref{sssec:API_GUI} subsection).
  \item Using \cmd{curl} or Postman to perform the HTTP requests directly.
\end{itemize}

\subsection{Grafana}
Once any device (along its entity) and particularly its subscription are properly created, CrateDB should start journaling the context data, therefore next step is to add CrateDB as a data source in Grafana.

@@

\begin{itemize}
  \item A rotary potentiometer does work well enough as a proof of concept, but it would have been more appropriate to have the expected linear potentiometer to try its sensitivity, which is a good point in order to measure such little variations as those produced in the stem diameter in a tree.
  \item The fact that the Raspberry Pi HAT uses a single channel chip (SX1276) has two important implications:
    \begin{itemize}
      \item In a first place, as indicated in its corresponding section, these kind of gateways are LoRaWAN compatible but not compliant. So it would be desirable to replace this HAT by another one, or simply to acquire a proper gateway for a production environment.
      \item Secondly, nodes cannot comunicate simultaneously with the gateway because it is listening to only one channel, so this is causing a time gap in data acquisition as it can be seen in picture \ref{fig:times_gap}
      \begin{figure}[ht]
        \centering
        \includegraphics[width=.9\textwidth]{../pictures/Grafana_times_gap.png}
        \caption{Small gap between times caused by the single channel chip.}
        \label{fig:times_gap}
      \end{figure}
    \end{itemize}
    \item End-nodes are being activated using ABP method, this method could not be as effective as OTAA method in case that downlink packets were transmitted. This is why due to some security implications, frame counters must be reset for ABP devices every time they are shutdown for any circumstance. In particular when downlink packets are transmitted, reset the frame counter in the devices could be necessary, so in this situation swapping to OTAA method might be worth considering. In fact, TTN offers the possibility to reset the frame counter in TTN console but this could be removed in the future.
    \item Maybe the developed GUI could have handled also the context data plots, however this task is left as an objective to be addressed in future researches, so this is not considered as part of the project because Grafana is completely capable to handle this for now and perhaps forever because it is extremely powerful handling databases.
    \item For a production environment it seems clear that Arduino devices should be encapsulated and the sensor (potentiometer) part must be improved; however this is not hard because of the wide offer for cases and power supply systems. This would lead to a very compact device which could be also improved adding more sensors due to CayenneLPP encoding standard's versatility which allows sending multiple measurements of different sensors. 
    \item Finally this project is a good candidate to replace expensive commercial and proprietary systems; it has potential to be developed further, depending on the needs of the researches.
\end{itemize}

>>>>

\begin{itemize}
  \item \textbf{PoE} (Power-over-Ethernet); this is not actually a suitable option because Raspberry Pi does not support PoE by itself, it needs a \href{https://www.raspberrypi.org/products/poe-hat/}{separate HAT} using the GPIO connector, so this makes it impossible to use it along with the LoRa HAT. Apparently there are hacks to reduce this HAT size, but they also use pins in the GPIO connector \cite{RaspiPoEHack}. Besides, this method only works with up to 100m cables, so it would be necessary to have a power source and internet access point within a 100m radius.\footnote{The \href{https://github.com/hallard/LoRasPI}{LoRasPI HAT} is more suitable for this option because does not cover whole GPIO connector, leaving free the GPIO pins used in \cite{RaspiPoEHack}.}
  \item \textbf{WiFi} and battery powered; this setup implies also the existence of a relatively close power source, because the access point used to provide that wireless network connection should work also with mains power. Probably the WiFi radio is not enough to make it the difference between this setup and the next one suggested.
  \item Simple \textbf{ethernet} and mains power, this setup is similar to the previous one, but omitting the WiFi limitations/configurations, the access point and the Raspberry Pi should be practically both at the same location.
  \item \textbf{GSM module} and battery or mains powered; this document considers this as the ideal setup. This setup does not require a conventional access point because the internet access is granted through a GSM module, in a similar way than a mobile phone. This would be the ideal way because it minimizes the resources needed, which is crucial in an environment where those probably will not be available ---in the middle of the forest.

@@

\begin{itemize}
  \item \textbf{IoTagent-LoRaWAN}, which is in charge of forwarding the traffic from TTN stack to Orion Context Broker.
  \item \textbf{Orion Context Broker}, this is the \myuline{core generic enabler}, in fact as indicated, this is the only one required component so the solution can be considered as \enquote{powered by FIWARE}. Besides, this generic enabler requires a \textbf{Mongo database} to store entities and subscriptions among other things. 
  \item \textbf{QuantumLeap}, in order to make the context data persistent, this generic enabler is needed. Orion Context Broker works mainly with two elements; entities and subscriptions, however, for any created entity the context data does not persist in Mongo database, instead the context data will be replaced/updated; so creating a subscription which notifies to QuantumLeap, this generic enabler will store the context data in a \textbf{Crate database} to make it persistent. Also a \textbf{Redis database} is used for geocoding purposes, however this project will not deal with it ---this is why of this database is not even being included in figure \ref{fig:GenView}.
  \item \textbf{Grafana}, this container is intended to provide a web interface to visualize and manage the retrieved context data. It stores its configuration and customization values in an embebed sqlite3 database; so there is not need for a separate container. 
\end{itemize}

The terms above highlighted in bold are the deployed containers that will be detailed later, many of them are not even part of FIWARE platform, though. First, this document will show how to set up the gateway, create an application and add devices inside The Things Network stack.

@@

\subsubsection{Grafana}\label{sssec:Grafana}
Grafana is a powerful dashboard used at many scenarios in the software industry, it allows monitoring almost everything the reader can imagine; because of this, it is capable of acquiring data from multiple sources. In this project it will be used in order to monitor the context data along the time.

@@

To get the panel plotting the context data stored in CrateDB, the reader must specify a query to retrieve the data from the database. This can be done in the \enquote{Query} tab below the graph shown in figure \ref{fig:panel_setup}: here the user could use the graphical interface or write the query manually by clicking on the pencil button at the right, so essentially the query should look as follows in example \ref{ej:CrateDB_query}

@@

Grafana has a wide options set intended to monitor different kind of systems, it can plot data in a variety of graphics and it even allows exporting to csv format, this can be useful for a more deep analysis using other software packages like, for instance, R, a well-known language widely used for statistical analysis.

Of course, this context data could have been acquired directly from CrateDB through some custom application developed specifically for this task, however this is beyond the purpose of this project. In a future, a feature like this could be implemented.

@@

When the requirements are satisfied and all docker containers are properly running\footnote{In a first place the GUI will retrieve the versions for Orion Context Broker, IoTagent-LoRaWAN and QuantumLeap, so if they are not running the GUI will not open.} the GUI can be run typing \cmd{./api-gui.py}; this should open the window shown in figure \ref{fig:GUI_app_keys}

@@

To check the Orion Context Broker, the IoTagent-LoRaWAN and the QuantumLeap versions running in the docker containers the reader has to click on the \enquote{About} tab, however, this first shown tab ---\enquote{Application Keys}--- is the most important one because without completing all the required information the \enquote{Devices Management} tab will not be enabled. Actually this tab performs a series of checks regarding TTN keys. Following the same TTN requirements,

@@

When the keys are updated properly the labels over the text fields are also updated with the chosen values and the \enquote{Devices Management} tab is enabled. This tab provides a user-friendly interface to manage devices. In a similar way, this info must be the same than the info provided by the TTN interface, so as in figure \ref{fig:device_management_tab}.

@@

This view allows the user to create and delete devices ---along with its entities and subscriptions--- nevertheless, the \enquote{Delete device and entities} button remains disabled until a device is chosen in the right list. In a similar way than \enquote{Applications Keys}, to add a new device to the FIWARE instance deployed by docker, all keys must fullfil a set of minimum requirements; as it can be seen in figure \ref{fig:adding_dev}

\begin{itemize}
  \item Device entity and ID must also be unique and it has to be at least 3 characters long. In fact, this ID must match the chosen ID in the TTN interface when the device is created.
  \item Application Session Key must be 32 characters long and also an hexadecimal string.
  \item Device EUI has to be 16 characters long and again an hexadecimal string.
\end{itemize}

These keys can be obtained from TTN as it can be seen in screenshot \ref{fig:dev_overview} ---this screenshot shows info about \cmd{node\_0}, though. We can see in figure \ref{fig:adding_dev}, that there was a problem with \enquote{Device EUI}, this was caused because the final \cmd{Z} letter in the string does not belong to hexadecimal characters set.

@@

Finally, as indicated, the \enquote{About} tab shows the versions info as it can be seen in figure \ref{fig:about_tab}. The QuantumLeap version is specified in \cmd{docker-compose.yml} file as well as the CrateDB version, this is to ensure compatibility between QuantumLeap and CrateDB; however, Orion Context Broker and IoTagent-LoRaWAN versions are not specified so they should always be the latests, this is why retrieving them could be helpful.

@@

\clearpage{\pagestyle{empty}\cleardoublepage}\thispagestyle{plain}  % Forces the next section to start in an odd page.
\section{Proof of concept, deployment}
Due to the unexpected circumstances related with the COVID19 pandemic during early 2020, this section is not as long as initially planned. Instead of testing the system in a real production environment, this section will focus on summarizing needed steps/stages to properly deploy the system. Although these steps had already been separately detailed, this section tries to put them together to give a consistent and practical point of view.

@@

\subsection{Set up the gateway}
This process has been widely detailed in the \ref{sssec:Gateway} subsection. The most important point in this stage is to ensure that the \cmd{systemd} service is working properly and to retrieve the Gateway ID, which will be needed in the next stage.

\subsection{TTN Stack}
After setting up the gateway, the reader should follow the three steps described in subsection \ref{sssec:TTN_Stack}: register a Gateway using its \enquote{Gateway ID}, add an application and register a device for that application. This is the first step because the reader will need the security keys for the two following stages:

\begin{itemize}
  \item Embed these keys in the Arduino sketch. This sketch will require the \enquote{Application Session Key}, \enquote{Network Session Key} and \enquote{Device Address}.
  \item Use them to create devices and subscriptions in FIWARE. To create the device ---and also its corresponding entity and subscription when using the designed graphical interface shown in the \ref{sssec:API_GUI} subsection--- the IoTagent will require the \enquote{Application ID}, \enquote{Application Access Key}, \enquote{Device ID}, \enquote{Application EUI} and again \enquote{Application Session Key}.
\end{itemize}

We must remember also that it is very important to choose the CayenneLPP as the payload decoder. So if these keys are available in the TTN platform, the process can continue.

\subsection{Customize the Arduino sketch for a specific device}
As indicated, this sketch will need the \enquote{Application Session Key}, the \enquote{Network Session Key} and \textbf{\enquote{Device Address}}. This latter is the most important one, because several nodes can use the same \enquote{Application Session Key} and \enquote{Network Session Key}, but the \enquote{Device Address} must be unique ---it is in fact also provided by TTN and the reader can check that every newly created node has a different one in TTN platform.

Of course, to be able to compile the sketch some dependencies must be satisfied, such as CayenneLPP, ArduinoJson and lmic-arduino libraries. Moreover, the \textbf{linear potentiometer has been replaced for a rotary potentiometer}, again due to the aforementioned circumstances in the moment when the project was developed.

@@

It is important to note that analog pin 2 has been chosen for analog signal acquisition. Besides, the reader can check the status of the device in the serial monitor, however, the reader may also note the serial monitor is setup at 115200 bits per second.

\subsection{Deploy Docker containers}
Once the hardware has been properly setup, the next step is to deploy locally all Docker containers, as indicated in the \ref{sssec:FIWARE} subsection. Using \cmd{docker-compose} this becomes an easy task.

This will deploy essentially the IoTagent-LoRaWAN, Orion CB, QuantumLeap and Grafana. The status of these containers can be checked typing \cmd{docker ps -a}.

\subsection{Devices, entities and subscriptions}
When Docker containers are deployed, it should be possible to send HTTP requests to create a device (along with its entity) and a subscription to notify QuantumLeap any change in any entity. This process can be done through two different ways as indicated previously,

\begin{itemize}
  \item Using the developed graphical user interface (recommended, see \ref{sssec:API_GUI} subsection).
  \item Using \cmd{curl} or Postman to perform the HTTP requests directly.
\end{itemize}

\subsection{Grafana}
Once any device (along with its entity) and particularly its subscription are properly created, CrateDB should start journaling the context data, therefore next step is to add CrateDB as a data source in Grafana.

@@

\begin{itemize}
  \item A rotary potentiometer does work well enough as a proof of concept, but it would have been more appropriate to have the expected linear potentiometer to try its sensitivity, which is a good point in order to measure such little variations as those produced in the stem diameter in a tree.
  \item The fact that the Raspberry Pi HAT uses a single channel chip (SX1276) has two important implications:
    \begin{itemize}
      \item In a first place, as indicated in its corresponding section, this kind of gateways are LoRaWAN compatible but not compliant. So it would be desirable to replace this HAT by another one, or simply to acquire a proper gateway for a production environment.
      \item Secondly, nodes cannot comunicate simultaneously with the gateway because it is listening to only one channel, so this is causing a time gap in data acquisition as it can be seen in picture \ref{fig:times_gap}
      \begin{figure}[ht]
        \centering
        \includegraphics[width=.9\textwidth]{../pictures/Grafana_times_gap.png}
        \caption{Small gap between times caused by the single channel chip.}
        \label{fig:times_gap}
      \end{figure}
    \end{itemize}
    \item End-nodes are being activated using ABP method, this method could not be as effective as OTAA method in case that downlink packets were transmitted. This is why due to some security implications, frame counters must be reset for ABP devices every time they are shutdown for any circumstance. In particular when downlink packets are transmitted, reset the frame counter in the devices could be necessary, so in this situation swapping to OTAA method might be worth considering. In fact, TTN offers the possibility to reset the frame counter in TTN console but this could be removed in the future.
    \item The developed GUI could have handled also the context data plots, however this task is left as an objective to be addressed in future developments, so this is not considered as part of the project because Grafana is completely capable to handle this for now and perhaps forever because it is extremely powerful handling databases.
    \item For a production environment it seems clear that Arduino devices should be encapsulated and the sensor (potentiometer) part must be improved; however this is not difficult because of the wide offer for cases and power supply systems. This would lead to a very compact device which could be also improved adding more sensors due to CayenneLPP encoding standard's versatility which allows sending multiple measurements of different sensors. 
    \item Finally this project is a good candidate to replace expensive commercial and proprietary systems; it has potential to be developed further, depending on the needs of the researchers.
\end{itemize}
